\DocumentMetadata{lang=en, pdfversion=1.7, pdfstandard=A-2u}
\documentclass[UKenglish]{uiomasterthesis}  %% ... or norsk or nynorsk or USenglish
\usepackage[utf8]{inputenc}                 %% ... or latin1
\usepackage[T1]{url}\urlstyle{sf}
\usepackage{babel, csquotes, graphicx, textcomp, frontpage/uiomasterfp, varioref}
\usepackage[backend=biber,style=numeric-comp]{biblatex}
\usepackage[hidelinks, hypertexnames=false]{hyperref}

% Custom packages
\usepackage{tcolorbox} % drawing boxes
\usepackage{glossaries} % acronyms
\usepackage{cleveref} % in-document references
\usepackage{siunitx} % \num
\usepackage{paralist} % inparenum
\usepackage{svg} % for SVG content on the front page


% Latex-related imports
\input{preamble/commands}
\input{preamble/acronyms}
\addbibresource{refs.bib}

% Set globals
\setglossarystyle{altlist}
\setdefaultenum{(1)}{(a)}{i.}{A.}
\interfootnotelinepenalty=10000 

\title{Can AI make our roads safer?}
\subtitle{Utilising Large language models to enhance Autonomous driving system test cases}
\author{Oliver Ruste Jahren}               

\begin{document}
\uiomasterfp[dept={Department of Informatics},
  program={Informatics: Programming and Systems architecture},
  supervisor={Shaukat Ali \and Karoline Nylænder},
  % or supervisors={A Name\and B Name},     %% if more than one
  info={Simula Research Laboratory},
  date={Fall 2025},
  long]

\frontmatter{}
\begin{abstract}
  Here come 3--6 sentences describing your thesis.
\end{abstract}

\begin{xabstract}[Sammendrag]
  Here comes the abstract in a different language.
\end{xabstract}

\tableofcontents{}
\listoffigures{}                            %% (omit if none)
% \listoftables{}                             %% (omit if none)

\begin{preface}
  Here comes your preface, including acknowledgments and thanks.
\end{preface}

\mainmatter{}
\part{Introduction}
\input{chapters/introduction}
\input{chapters/background}
\input{chapters/problem-description}
\input{chapters/literature-review}

\part{The project}
% \chapter{Planning the project} 

\input{chapters/related-work}
\input{chapters/proposal}
\chapter{Implementation architecture}
\chapter{Implementation details 1 - Thor}

Rest API - running test cases - Thor
“Fast API”, Python
POST a test case to the API. It will be ran on the ‘server’
RabbitMQ for listening for finished test cases? So that the client knows it can fetch the results?
Will need UUID for test cases so the correct result can be fetched after it has been ran
Need to store these somewhere. NoSQL database?
This component should also accumulate results.
Huge TODO: What metric are these results?
Should be containerised (Docker/Podman)

\chapter{Implementation details 2 - Odin}

Rest API - performing LLM enhancement - Odin
“Fast API”, Python
Take a base test case as body
Have some prompt repository
Apply prompts with LLMs
Must integrate with LLM. Either locally (Ollama) or remote (some API)
Look into good LLM agnostic transition layer. E.g. Aisuite
https://github.com/andrewyng/aisuite
Should use same UUIDs as outlined above, but suffixed with e.g. “pure” and “tainted”
Containerized. Docker compose?

\chapter{Implementation details 3 - Loki}

Client - orchestrating the process - Loki
Fetch available test cases from Thor? Select what/which are to be used
Store results clientside? Separate database for this?

\part{Conclusion}
\chapter{Experiment methodology}
\chapter{Results}
\chapter{Discussion}
\section{Environmental concerns}
Cost/benefit with using \acrshort{llms}. Refer back to \Nref{sec:llmProblems}.

\input{chapters/further-work}

\backmatter{}
\printbibliography{}
\end{document}
